### 2010-07-23 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: helplessly stare at whitespaces and parenthesis
Desc@ICFP: life and section 5 in colour^W color.
gaol: boiling the catamorphism

Our first attempt gets stuck quite quickly:
%
\[\stk{
\DescD : \Desc \\
\DescD \mapsto
  \tada{\stkm{\DUnit\\ \SYMBDSigma\\ \SYMBDIndx}}
       {\stkm{ \DUnit                                \\
         \DSigma{\Set}{(\LAM{\V{S}} \SHED)}      \\
         \DIndx{\DUnit}    }                    } \\
%\begin{array}{@{}ll}
%\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
%                               & \left[\begin{array}{l}
%                                   \DUnit                                \\
%                                   \DSigma{\Set}{\LAM{\V{S}} \SHED}      \\
%                                   \DIndx{\DUnit}                        \\
%                                 \end{array}
%                                 \right]
%\end{array}
}\]
%
Let us explain where we stand. Much as
we have done so far, we first offer a constructor choice from
$\DUnit$, $\SYMBDSigma$, and $\SYMBDIndx$. The reader will notice
that the `tagged' notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are defining.
For $\DUnit$, we immediately reach the end of the description. For
$\SYMBDIndx$, there is a single recursive argument. Describing
$\SYMBDSigma$ is problematic. Recall the specification of
$\SYMBDSigma$:
%
\[
\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
\]
%
So, we first pack a  $\Set$, $\V{S}$. We should then like
a recursive argument \emph{indexed} by $\V{S}$, but
that is an \emph{exponential}, and our presentation is entirely
first-order so far, delivering only sums-of-products. To code our
universe, we must first enlarge it!

@@

In Section~\ref{sec:desc-fix-point}, we hardwired a dependent
$\F{ind}$unction principle, instead of the catamorphism. However, in
some circumstances, the full power of a dependent elimination is not
necessary. Let us now derive the catamorphism
from $\F{ind}$ principle.

@@

The catamorphism is defined by induction on the description $\V{D}$,
with a readily propagated non-dependent return type $\V{T}$.
Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:
%
\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f} \mapsto
  \sind \LAM{\V{xs}}\LAM{\V{hs}} \V{f}\: \SHED
}\]
% 
We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{\Mu{\V{D}}}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). $\V{xs}$ is of the right shape, but its
sub-elements are of the wrong type. On the other hand, for each
sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding element
in $\V{T}$.  Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we must replace the recursive components of
\(\V{xs}\) by their counterparts from \(\V{hs}\). Let us write a
program to do that---please forgive us if we lapse to a pattern
matching notation, for readability's sake.
%
\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void   &
    \Void  \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}\: \V{d'}  &
    \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}\: \pair{\V{y}}{\V{d'}}{} &
    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}\: \pair{\V{g}}{\V{d'}}{} &
 \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
\end{array}
}\]
%
Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\:
(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the problem. In the
type theory, we have built a generic catamorphism. Any datatype will
now come equipped with this operation, for free.

@@

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a datatype specific
operation. This is a form of polytypic programming, as we learned from
PolyP~\cite{jansson:polyp}.

@@

{\small

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
Object                & Role                        & Status \\
\hline
\hline
$\EnumU$              & Build finite sets           & Levitated \\
\hline
$\Desc$               & Describe pattern functors   & Levitated \\
\hline
$\SYMBdescop{\_}$     & Interpret descriptions      & Hardwired \\
\hline
$\SYMBMu$, $\SYMBCon$ & Define, inhabit fixpoints   & Hardwired \\
\hline
$\SYMBind$, $\SYMBAll$, $\SYMBall$  
                      & Induction principle         & Hardwired \\
\hline
\end{tabular}
\end{center}
}

>>>>

Our first attempt gets stuck quite quickly:
%
\[\stk{
\DescD : \Desc \\
\DescD \mapsto
  \tada{\stkm{\DUnit\\ \SYMBDSigma\\ \SYMBDIndx}}
       {\stkm{ \DUnit                                \\
         \DSigma{\Set}{\LAM{\V{S}} \SHED}      \\
         \DIndx{\DUnit}    }                    } \\
%\begin{array}{@{}ll}
%\DescD \mapsto \DSigma{\!}{\!} & (\EnumT{[ \DUnit, \DSigma{\!}{\!}, \DIndx{\!} ]})  \\
%                               & \left[\begin{array}{l}
%                                   \DUnit                                \\
%                                   \DSigma{\Set}{\LAM{\V{S}} \SHED}      \\
%                                   \DIndx{\DUnit}                        \\
%                                 \end{array}
%                                 \right]
%\end{array}
}\]
%
Let us explain where we stand. Much as
we have done so far, we first offer a constructor choice from
$\DUnit$, $\SYMBDSigma$, and $\SYMBDIndx$. The reader will notice
that the `tagged' notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are defining.
For $\DUnit$, we immediately reach the end of the description. For
$\SYMBDIndx$, there is a single recursive argument. Describing
$\SYMBDSigma$ is problematic. Recall the specification of
$\SYMBDSigma$:
%
\[
\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
\]
%
So, we first pack a  $\Set$, $\V{S}$. We should then like
a recursive argument \emph{indexed} by $\V{S}$, but
that is an \emph{exponential}, and our presentation is entirely
first-order so far, delivering only sums-of-products. To code our
universe, we must first enlarge it!

@@

In Section~\ref{sec:desc-fix-point}, we hardwired a dependent
$\F{ind}$uction principle, but sometimes,
iteration suffices. Let us construct the catamorphism.

@@

We proceed by induction on the \emph{data} in \(\Mu{\V{D}}\):
the non-dependent return type $\V{T}$ is readily propagated.
Given a node $\V{xs}$
and the induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:
%
\[\stk{
\cata : \PITEL{\V{D}}{\Desc}
           \PI{\V{T}}{\Set}
           (\descop{\V{D}}{\V{T}} \To \V{T}) \To 
           \Mu{\V{D}} \To \V{T} \\
\cata\: \V{D}\: \V{T}\: \V{f} \mapsto
  \sind \LAM{\V{xs}}\LAM{\V{hs}} \V{f}\: \SHED
}\]
% 
The hole remains: we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{\Mu{\V{D}}}{(\LAM{\_} \V{T})}{\V{xs}}}\)
to hand, and we need a \(\descop{\V{D}}{\V{T}}\).
Now, $\V{xs}$ has the right shape, but its
components have the wrong type. However, for each such
component, $\V{hs}$ holds the corresponding value
in $\V{T}$.  We need a function to \(\F{replace}\) the former with the latter:
this pattern
matching sketch yields an induction on \(\V{D}\). We fill the hole
with \(\F{replace}\: \V{D}\:(\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\).
%
\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\begin{array}{@{}l@{\:}l@{\:}l@{\:\:\mapsto\:\:}l}
\F{replace}\: \DUnit &          \V{X}\: \V{Y}\: \Void&          \Void   &
    \Void  \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})& \V{X}\: \V{Y}\: \pair{\V{s}}{\V{d}}{}& \V{d'}  &
    \pair{\V{s}}{\F{replace}\: (\V{D}\: \V{s})\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
\F{replace}\: (\DIndx{\V{D}})&     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{d}}{}& \pair{\V{y}}{\V{d'}}{} &
    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{} \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})& \V{X}\: \V{Y}\: \pair{\V{f}}{\V{d}}{}& \pair{\V{g}}{\V{d'}}{} &
 \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{d}\: \V{d'}}{}
\end{array}
}\]
%

@@

We have shown how to derive a generic
operation, \(\cata\), from a pre-existing generic operation,
\(\F{ind}\), by
manipulating descriptions as data: the catamorphism is just a
function taking each $\Desc$ value to a datatype specific
operation. This is polytypic programming, as in
PolyP~\cite{jansson:polyp}, made ordinary.

@@

{
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
Object                & Role                        & Status \\
\hline
\hline
$\EnumU$              & Build finite sets           & Levitated \\
\hline
$\Desc$               & Describe pattern functors   & Levitated \\
\hline
$\SYMBdescop{\_}$     & Interpret descriptions      & Hardwired \\
\hline
$\SYMBMu$, $\SYMBCon$ & Define, inhabit fixpoints   & Hardwired \\
\hline
$\SYMBind$, $\SYMBAll$, $\SYMBall$  
                      & Induction principle         & Hardwired \\
\hline
\end{tabular}
\end{center}
}
